---
title: "Final Project"
author: "Juyee Sabade"
date: "4/3/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(patchwork)
library(fpp3)
options(digits=2)
library(tseries)
library(readxl)
library(dplyr)
library(pacman)
library(tsibble)
library(tidyverse)

p_load(data.table,lubridate,collapse,ggplot2,fixest,stringr,modelsummary,eeptools)
```


## Importing data

```{r}
sanfranscico <- fread("..//Data//formatted_sanfrancisco.csv")
```


```{r}
setDT(sanfranscico)
avg_times <- sanfranscico[,mean_travel_time:= mean(MeanTravelTimeSeconds),by=Date]
```


# Time Series Analysis & Visualizations

## Analyzing Mean travel times of daily Uber trips in San Francisco

```{r}
avg_times <- avg_times %>% select(Date,mean_travel_time) %>% distinct()
```

```{r}
avg_times <- avg_times %>% mutate(day=weekdays(Date))
```


```{r}
ggplot(avg_times)+
  geom_point(aes(x=Date,y=mean_travel_time),color="cornflowerblue", size = 2, alpha=.8)+
  labs(title="Time plot of mean travel duration")
```

As we can observe here, the mean travel time decreases around the end of the year. And we can also see a very steep and uncharacteristic drop after February 2020 which is due to the effects of the pandemic. For the purpose of or project, we will only take data till January 2020 as the pandemic is a one time occurrence which can add bias to our forecasting. 

```{r}
#Slicing the last 2 months of data
avg_times <- slice(avg_times,1:(n()-60))
```

```{r}
ggplot(avg_times)+
  geom_point(aes(x=Date,y=mean_travel_time),color="cornflowerblue", size = 2, alpha=.8)+
  labs(title="Time plot of mean travel duration before the pandemic")
```

Here we can observe the mean travel time decreases around the end of each year and remains less till January of the new year.

## Summary statistics and Histogram

```{r}
summary(avg_times)
```

On an average it takes 1089 seconds to go from the central part of San Francisco to any other region. The minimum duration is 602 which was in 2016 and the maximum duration is 1439. This could mean that traffic has increased ovwer the years or there are a larger number of trips to zone farther out from the origin. Also this indiates that our data is left skewed as there are a longer concentration of longer trips than shorter trips.
We can confirm that from the below histogram as well.


```{r}
hist(avg_times$mean_travel_time)
```



## Analyzing total number of Uber trips in San Francisco everyday from the central point of the city to all the destination zones.


```{r}
trip_count <- sanfranscico[,count_trips:= .N,by=Date] 
trip_count <- trip_count %>% select(Date,count_trips) %>% distinct()
#Slicing last two months of data to exclude pandemic bias
trip_count <- slice(trip_count,1:(n()-60))
```


```{r}
trip_count <- trip_count %>% mutate(day=weekdays(Date))
```


```{r}
ggplot(trip_count)+
  geom_point(aes(x=Date,y=count_trips),color="cornflowerblue", size = 2, alpha=.8)+
  labs(title = "Time plot of trip counts before pandemic")
```


```{r}
merged_data <- merge(avg_times,trip_count,by="Date")
```

```{r}
ggplot(merged_data) +
  geom_point(aes(x=count_trips,y=mean_travel_time),color="steelblue") +
  labs(x="Number of Destinations",y="Mean Travel Time (in seconds)",title="Correlation between travel time and number of destinations")
```

As we can observe in the above plot, there is a linear relationship between number of destination zones and mean travel time because a higher number of trips could indicate more traffic which could in turn lead to higher travel times.

## Calculating weekly demand statistics

```{r}
week_stats <- sanfranscico %>% group_by(DayOfWeek) %>% summarise(count_trips = n())
ggplot(week_stats,aes(x=DayOfWeek,y=count_trips,fill=DayOfWeek,group=DayOfWeek))+
  geom_bar(stat="identity")+
  labs(title="Number of trips on each day of the week",x="Day of the Week",y="Number of trips")
```

We see the highest number of trips on weekends. There is not much difference among the other weekdays.


```{r}
ggplot(avg_times,aes(x=Date,y=mean_travel_time,color=day), size = 2, alpha=.8)+
  geom_point()+
  labs("Time plot of mean travel duration")
```


## Analyzing mean travel times faceted out by weekdays

```{r}
ggplot(avg_times)+
  geom_point(aes(x=Date,y=mean_travel_time),color="cornflowerblue", size = 2, alpha=.8)+
  facet_wrap(~day, scales="free_y") +
  labs(title="Time plot of mean travel time")
```

Here we observe that the mean travel time is lesser on Mondays, Fridays and Sundays. That might be either because there are lesser Uber trips on those days or there is lesser traffic on the road on those days. Around the year 2020, the mean travel time starts reducing on all the days.


## Analyzing daily trip counts facted out by weekdays

```{r}
ggplot(trip_count)+
  geom_point(aes(x=Date,y=count_trips),color="cornflowerblue", size = 2, alpha=.8)+
  facet_wrap(~day, scales="free_y") +
  labs(title="Time plot of trip counts")
```

Here we observe that the number of trips are usually highest on weekends and lesser on weekdays. This could be because lot of people might be 
going out on weekend.


```{r}
avg_times <-  avg_times %>% mutate(Date_parsed = date(Date)) %>% select(-Date)
avg_times_ts <- as_tsibble(avg_times,index=Date_parsed)
```


```{r}
avg_times_ts <- fill_gaps(avg_times_ts,mean_travel_time = mean(mean_travel_time))
```

## Performing Weekly and Yearly Seasonality

```{r}
avg_times_ts %>% gg_season(mean_travel_time) + labs(y="Mean Travel Time",x="Month",title="Yearly Seasonal plot of mean travel times")
```

We are seeing seasonal variations wherein there is a decrease in January and December in mean travel times.


```{r}
avg_times_ts %>% gg_season(mean_travel_time,period="week") + labs(y="Mean Travel Time",x="Day",title="Weekly Seasonal plot showing daily seasonal patterns of mean travel times")
```



```{r}
trip_count <-  trip_count %>% mutate(Date_parsed = date(Date)) %>% select(-Date)
trip_count_ts <- tsibble(trip_count,index=Date_parsed)
trip_count_ts <- fill_gaps(trip_count_ts,.full = TRUE)
```



```{r}
trip_count_ts <- fill_gaps(trip_count_ts,.full = TRUE)
```


```{r}
trip_count_ts %>% gg_season(count_trips) + labs(y="Number of Trips",x="Month",title="Yearly seasonal plot of number of destination zones")
```




```{r}
trip_count_ts %>% gg_season(count_trips,period="week") + labs(y="Number of Trips",x="Day",title="Weekly Seasonal plot of number of destination zones")
```

## Lag Plots and ACF plots


```{r}
avg_times_ts %>% gg_lag(mean_travel_time,geom="point")+
  labs(x="lag(Mean Travel Time)",title="Lag Plot of Mean Travel times")
```




```{r}
avg_times_ts %>% ACF(mean_travel_time,lag_max=48) %>% autoplot() +
  labs(title="ACF Plot of Mean Travel Times")
```

Here we are seeing a scalloped shape because there is a weekly seasonality that can be observed. The strongest correlation can be seen among the starting two consecutive days.


```{r}
trip_count_ts %>% ACF(count_trips,lag_max=48) %>% autoplot() +
  labs(title="ACF Plot of Number of Trips")
```

Even in this ACF plot, we are observing a very strong weekly seasonality in the data. So we can confirm that the number of trips on a certain day of the week are fairly predictable. 


## Plotting 30 day moving average

```{r}
avg_times_ma <- avg_times_ts %>% mutate(
    `30-MA` = slider::slide_dbl(mean_travel_time, mean,
                .before = 15, .after = 14, .complete = TRUE)
  )
```


```{r}
ggplot(avg_times_ma) +
  geom_line(aes(x=Date_parsed,y = `30-MA`), colour = "#D55E00") +
  labs(y = "Mean Travel Time",
       title = "30 day moving average of travel time in seconds") +
  guides(colour = guide_legend(title = "series"))
```

Here we can see that the seasonality pattern that we identified in the previous plots wherein the mean travel time reduces around the end of each year and remains lesser till January is more clearly identifiable.


```{r}
trip_counts_ma <- trip_count_ts %>% mutate(
    `30-MA` = slider::slide_dbl(count_trips, mean,
                .before = 15, .after = 14, .complete = TRUE)
  )
```


```{r}
ggplot(trip_counts_ma) +
  geom_line(aes(x=Date_parsed,y = `30-MA`), colour = "#D55E00") +
  labs(y = "Number of Trips",
       title = "30 day moving average of trip counts") +
  guides(colour = guide_legend(title = "series"))
```

We can observe the same seasonality pattern in this plot as well which also confirms that trip counts are correlated with mean travel time. 


```{r}
lambda <- avg_times_ts %>%
  features(mean_travel_time, features = guerrero) %>%
  pull(lambda_guerrero)

avg_times_ts %>%
  autoplot(box_cox(mean_travel_time, lambda)) +
  labs(y = "",
       title = latex2exp::TeX(paste0(
         "Transformed mean travel times with $\\lambda$ = ",
         round(lambda,2))))
```

## Calculating Decomposition Features

```{r}
avg_times_ts %>% features(mean_travel_time,feat_stl)
```


```{r}
trip_count_ts %>% features(count_trips,feat_stl)
```


# Time series decompositon of mean travel time

## Classical multiplicative Decomposition

```{r}
dcmp <- avg_times_ts %>%  model(classical_decomposition(mean_travel_time,type="multiplicative")) 

components(dcmp) %>% autoplot()
```

Here we are not seeing a very good decompositon as there is still a yearly seasonal component left in our trend component.

## STL Decomposition

```{r}
library(feasts)
 
dcmp <- avg_times_ts %>% model(stl=STL(mean_travel_time ~ trend(window=31) +
                                          season(window="periodic"),
                                        robust=TRUE)) 

components(dcmp) %>% autoplot()
```

```{r}
dcmp <- avg_times_ts %>% model(stlf=decomposition_model(STL(log(mean_travel_time) ~ trend(window=30),
                                        robust=TRUE),
                               SNAIVE(season_adjust)))

dcmp %>% forecast(h=90) %>% autoplot(avg_times_ts)
```

Here we are observing a good time series decomposition because there is a seperate seasonal component for yearly seasonality and weekly seasonality.


```{r}
avg_times_ts %>% model(SNAIVE(log(mean_travel_time))) %>% forecast(h=365) %>% autoplot(avg_times_ts) 
```


```{r}
dcmp %>% gg_tsresiduals()
```

It looks like there is very little information left in the residuals.


## Running STL decomposition on a training set 

```{r}
train <- slice(avg_times,1:(n()-60))
train <- as_tsibble(train,index=Date_parsed)
train <- fill_gaps(train,mean_travel_time = mean(mean_travel_time))
```

```{r}
train_dcmp <- train %>% model(stlf=decomposition_model(STL(log(mean_travel_time) ~ trend(window=31),
                                        robust=TRUE),
                               NAIVE(season_adjust)))

```


```{r}
train_dcmp %>% forecast(h=60) %>% autoplot(avg_times_ts,level=NULL,color="pink")
```

```{r}
train_fc <- train_dcmp %>% forecast(h=60)
accuracy(train_fc,avg_times_ts)
```

## Forecasting with STL decomposition

```{r}
fit_dcmp_new <- avg_times_ts %>%
  model(stlf = decomposition_model(
    STL(mean_travel_time ~ trend(window = 7), robust = TRUE),
    NAIVE(season_adjust)))
```


```{r}
fit_dcmp_new %>%
  forecast(h=90) %>%
  autoplot(avg_times_ts)+
  labs(y = "Mean travel time",
       title = "Forecasting Mean travel times")
```

```{r}
predictions <- fit_dcmp_new %>%
  forecast(h=90) %>% hilo()
```

# ETS Modelling


```{r}
fit_ets <- train %>% model(additive = ETS(mean_travel_time ~ error("A") + trend("A") +
                                                season("A")),
    hw = ETS(mean_travel_time ~ error("M") + trend("Ad") +
                                                season("M")))

fit_ets_fc <- fit_ets %>% forecast(h=60)
```

```{r}
fit_ets_fc %>% autoplot(avg_times_ts,level=NULL)
```

## Performance Statistics

```{r}
report(fit_ets)
```

From the AIC and BIC values, it can be observed that ETS is not performing well. 

## Decomposition

```{r}
components(fit_ets) %>% autoplot()
```

# Transformations for weekly and monthly data


```{r}
avg_times <- avg_times %>% mutate(month=yearmonth(Date_parsed))

avg_times_mo <- avg_times %>% group_by(month) %>% summarise(mean_travel_time = mean(mean_travel_time))

avg_times_mo <- as_tsibble(avg_times_mo,index=month)

avg_times_mo <- fill_gaps(avg_times_mo,mean_travel_time = mean(mean_travel_time))


avg_times <- avg_times %>% mutate(week=yearweek(Date_parsed))

avg_times_week <- avg_times %>% group_by(week) %>% summarise(mean_travel_time = mean(mean_travel_time))

avg_times_week <- as_tsibble(avg_times_week,index=week)

avg_times_week <- fill_gaps(avg_times_week,mean_travel_time = mean(mean_travel_time))

```


```{r}
autoplot(avg_times_week)
```



# Stationarity and differencing

```{r}
avg_times_week %>% features(log(mean_travel_time),unitroot_nsdiffs)
```

Here we can observe through unit root test that the weekly transformed data requires one order of seasonal differencing.

```{r}
avg_times_week  %>% mutate(log_travel_time =difference(log(mean_travel_time),52)) %>% features(log_travel_time,unitroot_ndiffs)
```


```{r}
avg_times_week %>% mutate(doubly_diff_travel_time = difference(difference(log(mean_travel_time),52))) %>%
  features(doubly_diff_travel_time,unitroot_kpss)
```
Here we can see that the KPSS test is giving a high p value which means our first order differencing and first order seasonal differencing is correct.

```{r}
avg_times_week <- avg_times_week %>% mutate(log_travel_time = difference(log(mean_travel_time),52))

autoplot(avg_times_week,log_travel_time)
```

```{r}
avg_times_week <- avg_times_week %>% mutate(doubly_diff_travel_time = difference(difference(log(mean_travel_time),52)))
```


```{r}
autoplot(avg_times_week,doubly_diff_travel_time)
```


# ARIMA Modelling

```{r}
avg_times_week %>% gg_tsdisplay(doubly_diff_travel_time,plot_type = "partial",lag=52)
```


Here we are seeing some minor spikes. The significant spike at lag 1 in the ACF suggests a non-seasonal MA(1) component. The significant spike at lag 52 in the ACF suggests a seasonal MA(1) component. Consequently, we begin with an ARIMA(0,1,1)(0,1,1)52 model, indicating a first difference, a seasonal difference, and non-seasonal MA(1) and seasonal MA(1) component. If we had started with the PACF, we may have selected an ARIMA(3,1,0)(0,1,1)52 model — using the PACF to select the non-seasonal part of the model and the ACF to select the seasonal part of the model. We will also include an automatically selected model.

```{r}
fit <- avg_times_week %>%
  model(arima011011 = ARIMA(mean_travel_time ~ pdq(0,1,1) + PDQ(0,1,1)),
    arima310010 = ARIMA(mean_travel_time ~ pdq(3,1,0) + PDQ(0,1,1)),
    auto = ARIMA(mean_travel_time, stepwise = FALSE, approx = FALSE))
```


```{r}
res <- fit %>% pivot_longer(everything(), names_to = "Model name",
                     values_to = "Orders")
```

```{r}
res
```


```{r}
glance(fit) %>% arrange(AICc) %>% select(.model:BIC)
```


We can observe that the ARIMA(0,1,1)(0,1,1)52 model we predicted is performing the best with the lowest AIC. We can further confirm it by looking at its residual diagnostics.

```{r}
fit %>% select(arima011011) %>% gg_tsresiduals(lag=52)
```
We can see that there is no information left in the residuals, so the model is performing well.

```{r}
augment(fit) %>% filter(.model == "arima011011") %>%
  features(.innov, ljung_box, lag=52, dof=2)
```
On performing Ljung-Box test, we get a high p-value which confirms our residuals are white noise.

```{r}
augment(fit) %>% filter(.model == "auto") %>%
  features(.innov, ljung_box, lag=52, dof=6)
```


## ARIMA Forecasting

We will perform ARIMA forecasting of mean travel times with the best fitted model.

```{r}
forecast(fit,h=365) %>%
  filter(.model=='arima011011') %>% autoplot(avg_times_week) +
  labs(title = "Forecasting mean travel times",
       y="Mean travel time")

```

We can see that the ARIMA model is able to correctly capture the yearly seasonality. With respect to trend, its predicting high traffic flow which means high Uber demand in the coming years.


# Dynamic Harmonic Progression

When there are long seasonal periods, a dynamic regression with Fourier terms is often better than other models. Seasonal pattern is modelled using Fourier terms with short-term time series dynamics handled by an ARMA error.

```{r}
fit_harmonic <- avg_times_ts %>% model(
  dhr = ARIMA(sqrt(mean_travel_time) ~ PDQ(0,0,0) + pdq(d=0)+
    fourier(period=7,K=3)+
    fourier(period=30,K=2))
)

fc_fit_harmonic <- fit_harmonic %>% forecast(h=365)
```


```{r}
fc_fit_harmonic %>% autoplot(avg_times_ts)
```

```{r}
fit_hr <- model(avg_times_week,
                `K = 7` = ARIMA(log(mean_travel_time) ~ fourier(K=7) + PDQ(0,0,0)),
                `K = 8` = ARIMA(log(mean_travel_time) ~ fourier(K=8) + PDQ(0,0,0)),
                `K = 9` = ARIMA(log(mean_travel_time) ~ fourier(K=9) + PDQ(0,0,0)),
                `K = 10` = ARIMA(log(mean_travel_time) ~ fourier(K=10) + PDQ(0,0,0)),
                `K = 11` = ARIMA(log(mean_travel_time) ~ fourier(K=11) + PDQ(0,0,0)),
                `K = 12` = ARIMA(log(mean_travel_time) ~ fourier(K=12) + PDQ(0,0,0))
                 )
```


```{r}
fit_hr %>% forecast(h="2 years") %>% autoplot(avg_times_week,level=95) + 
  facet_wrap(vars(.model),ncol=2)+
  guides(colour="none",fill="none",level="none")+
  geom_label(
    aes(x = yearweek("2015 W53"), y = 1500,
        label = paste0("AICc = ", format(AICc))),
    data = glance(fit_hr))+
  labs(title= "Mean Travel times forecast",
       y="Mean travel time")
```

The AIC value is the least for K=12. The forecasts are also able to capture the seasonality.



# Prophet Modelling

```{r}
#install.packages("fable.prophet")
library(fable.prophet)
```


```{r}
train_wk <- slice(avg_times_week,1:(n()-60))
#train <- as_tsibble(train,index=Date_parsed)
#train <- fill_gaps(train,mean_travel_time = mean(mean_travel_time))
```



```{r}
fit_prop <- train_wk %>% model (
  arima = ARIMA(mean_travel_time),
  prophet = prophet(mean_travel_time ~ season(period=52,order=10,type="multiplicative"))
)

fc_prop <- fit_prop %>% forecast(h=60)
```

```{r}
fc_prop %>% autoplot(avg_times_week)
```

```{r}
fc_prop %>% accuracy(avg_times_week)
```


## Prophet Decomposition & Forecasting 

```{r}
fit_ph_dcmp <- avg_times_week %>% model(
  prophet(log(mean_travel_time) ~
            season(period = "month", order = 5) +
            season(period="year",order=10))
)

fit_ph_dcmp %>% 
  components() %>%
  autoplot()

```

```{r}
fit_ph_dcmp %>% gg_tsresiduals()
```

## Prophet modelling on 90% training data

```{r}
test_rows = as.integer(((100-90)/100)*nrow(avg_times_week))
train <- slice(avg_times_week,1:(n()-test_rows))
#train <- as_tsibble(train,index=Date_parsed)
train <- fill_gaps(train,mean_travel_time = mean(mean_travel_time))
train   
```


```{r}
fit_ph_dcmp <- train %>% model(
  prophet(mean_travel_time ~
            season(period = "month", order = 5) +
            season(period="year",order=10))
)

fit_ph_dcmp %>% 
  components() %>%
  autoplot()
```


```{r}
fc_ph <- fit_ph_dcmp %>% forecast(h=60) 

fc_ph %>% autoplot(avg_times_week)

```



```{r}
metrics <- fc_ph %>% 
          accuracy(avg_times_week) 
        
metrics %>% pivot_longer(!c(.model,.type),names_to="features",values_to="value") %>%
          select(features,value)
```

We can observe that Prophet is doing a decent job at predicting values with a 5.13% mean absolute percentage error from the actual values.


Now we can forecast future mean travel times with Prophet.

```{r}
fit_ph_dcmp <- avg_times_week %>% model(
  prophet(log(mean_travel_time) ~
            season(period = "month", order = 5) +
            season(period="year",order=10))
)

fit_ph_dcmp %>% 
  components() %>%
  autoplot()
```
```{r}
fit_ph_dcmp %>% gg_tsresiduals()
```
```{r}
fc_ph <- fit_ph_dcmp %>% forecast(h=208) 

fc_ph %>% autoplot(avg_times_week)

```

Interestingly, here we can observe that Prophet is predicting a downward trend in the coming years as opposed to what ARIMA predicted. Prophet is also able to capture the correct yearly seasonality.


# Model Comparison - ARIMA, ETL, Prophet

```{r}
train <- slice(avg_times_week,1:(n()-60))
train <- as_tsibble(train,index=week)
train <- fill_gaps(train,mean_travel_time = mean(mean_travel_time))
```


```{r}
fit_compare <- train %>%
  model(
    arima = ARIMA(mean_travel_time),
    ets = ETS(mean_travel_time),
    prophet = prophet(log(mean_travel_time) ~ season(period = "month", order = 5) +
            season(period="year",order=10))
  )
```

```{r}
fc <- fit_compare %>% forecast(h=60)
fc %>% autoplot(avg_times_week)
```

```{r}
fc %>% accuracy(avg_times_week)
```

Here we can see that ARIMA is performing better than the other models with the lowest RMSE.
